Machine learning topological phases in real space

### 2020-07-25 08_arxiv/article.tex
Generated new arxiv submission

\definecolor{antiquefuchsia}{rgb}{0.57, 0.36, 0.51}

@@

\begin{abstract}
We develop a supervised machine learning algorithm that is able to learn topological phases of finite condensed matter systems from bulk data in real lattice space. The algorithm employs diagonalization in real space together with any supervised learning algorithm to learn topological phases through an eigenvector ensembling procedure. We combine our algorithm with decision trees and random forests to successfully recover topological phase diagrams of Su-Schrieffer-Heeger (SSH) models from bulk lattice data in real space and show how the Shannon information entropy of ensembles of lattice eigenvectors can be used to retrieve a signal detailing how topological information is distributed in the bulk. We further use insights obtained from these information entropy signatures to engineer global topological features from real space lattice data that still carry most of the topological information in the lattice, while greatly diminishing the size of feature space, thus effectively amounting to a topological lattice compression. Finally, we explore the theoretical possibility of interpreting the information entropy topological signatures in terms of emergent information entropy wave functions, which allows us to establish Heisenberg and Hirschman uncertainty relations for topological phase transitions. The discovery of Shannon information entropy signals associated with topological phase transitions from the analysis of data from several thousand SSH systems illustrates how model explainability in machine learning can advance the research of exotic quantum materials with properties that may power future technological applications such as qubit engineering for quantum computing. 
\end{abstract}

@@

The advantage of using decision tree-based algorithms to learn topological phases from local eigenvector data is that their use of entropy-based cost functions (such as Shannon information entropy or Gini impurity) furnishes them with an intrinsinc model explanaibility tool that summarizes how important each feature was to learn the desired patterns in the data. This makes it much easier to trace the localization of relevant information along the features of a data set. Here we use the Shannon information entropy of ensembles of real space eigenvectors to recover a signal quantifying the amount of topological information available from each lattice site. This is a highly non-trivial proposition since the topological phase of a system is a global property of the system as a whole emerging from complex interactions between its components, and therefore even defining a local topological signal is a daunting theoretical task. To our knowledge this is the first time that a signal describing the localization of topological information in the bulk of topological condensed matter systems is presented in the literature. These topological signals, here called information entropy signatures, are the subject of section \ref{information_entropy_signatures}.

@@

As of yet model explainability \cite{gilpin2018explaining}-\cite{roscher2020explainable} is one of the topics at the edge of machine learning research that has been little explored by the physics community working at the interface between the two disciplines. This raises important questions as to whether machine learning can in fact help to advance theoretical investigation in physics, since the majority of physics papers published on the subject are proofs of concept aimed at showing that modern machine learning techniques are capable of recognizing the relevant patterns in data from physical systems whose properties were known in advance. By proposing new concepts from the data analysis of physical systems of contemporary interest and knitting together ideas from topological phase transitions and information theory by dint of model explainability, we expect to draw the physics community's attention to this essential machine learning tool. 
\begin{figure}
\centering
\subfigure[]{\label{ssh1}\includegraphics[width=.45\textwidth]{./phase_diagrams/ssh1.png}}\quad
\subfigure[]{\label{ssh2}\includegraphics[width=.45\textwidth]{./phase_diagrams/ssh2.png}}
\caption{Phase diagrams in parameter space. a) SSH model with first-neighbor hoppings $t_1$ and $t_2$. The (red) regions with winding number $W$ = 0 are trivial, while the (blue) regions with winding number $W$ = 1 are topologically non-trivial. b) SSH model with first ($t_1$ and $t_2$) and second ($T_1$ and $T_2$) nearest-neighbor hoppings. In this article we set $t_1$ = $t_2$ = 1 and renamed the variables $T_1$ $\rightarrow$ $t_1$, $T_2$ $\rightarrow$ $t_2$ for convenience. The (orange) region with winding number $W$ = 0 is trivial while the others with winding numbers $W$ = -1, $W$ = 1 and $W$ = 2 (red, green and blue respectively) are topologically non-trivial.}
\label{fig:phasediagrams}
\end{figure}

@@

Figures \ref{figexp1_exp} and \ref{figexp2_exp} respectively illustrate single iterations of experiment 1 (section \ref{experiment_1}) and experiment 2 (section \ref{experiment_2}) as seen from parameter space. The accuracy statistics presented in the following subsections, as well as the probability heatmaps and recovered phase diagrams shown in figures \ref{ssh1_heatmaps} and \ref{ssh2_heatmaps} were obtained after bootstrapping each experiment $n_{exp}$ = \nExp\ times. Thus, each probability heatmap shown in figures \ref{ssh1_heatmap_0}, \ref{ssh1_heatmap_1} and \ref{ssh2_heatmap_-1} to \ref{ssh2_heatmap_2} represents the averaged fraction of eigenvectors of each Hamiltonian in the grid that were classified with a given winding number across $n_{exp}$ experiments. The recovered phase diagrams \ref{ssh1_heatmap} and \ref{ssh2_heatmap} are constructed by superposing the correponding probability heatmaps. As these figure make clear, the recovered phase diagrams faithfully portray the true phase diagrams in figure \ref{fig:phasediagrams}, with clear phase transition lines appearing in the regions of highest uncertainty.

@@

\begin{figure}
\centering
\subfigure[]{\label{figexp1_exp:a}\includegraphics[width=.32\textwidth]{./ssh1/real_space_all_sites/scatter_train_val_test_experiment_0.png}}\quad
\subfigure[]{\label{figexp1_exp:b}\includegraphics[width=.32\textwidth]{./ssh1/real_space_all_sites/scatter_winding_train_experiment_0.png}}
\subfigure[]{\label{figexp1_exp:c}\includegraphics[width=.32\textwidth]{./ssh1/real_space_all_sites/pcolormesh_prediction_grid_experiment_0.png}}
\caption{Visualization of a single iteration of experiment 1 (section \ref{experiment_1}) as seen from 2D parameter space. (a) Train/validation/test split. (b) Distribution of winding numbers in the training set. (c) Phase diagram learned from real space lattice data by a combination of decision tree and eigenvector ensembling.}
\label{figexp1_exp}
\end{figure}

@@

\begin{figure}
\centering
\subfigure[]{\label{figexp2_exp:a}\includegraphics[width=.32\textwidth]{./ssh2/real_space_all_sites/scatter_train_val_test_experiment_0.png}}\quad
\subfigure[]{\label{figexp2_exp:b}\includegraphics[width=.32\textwidth]{./ssh2/real_space_all_sites/scatter_winding_train_experiment_0.png}}
\subfigure[]{\label{figexp2_exp:c}\includegraphics[width=.32\textwidth]{./ssh2/real_space_all_sites/pcolormesh_prediction_grid_experiment_0.png}}
\caption{Visualization of a single iteration of experiment 2 (section \ref{experiment_2}) as seen from 2D parameter space. (a) Train/validation/test split. (b) Distribution of winding numbers in the training set. (c) Phase diagram learned from real space lattice data by a combination of random forest and eigenvector ensembling.}
\label{figexp2_exp}
\end{figure}

@@

We should briefly comment on the possibility of using the Gini impurity of ensembles of eigenvectors \cite{friedman2001elements} instead of their Shannon entropy. This would similarly lead us to Gini impurity signatures of topological phase transitions. Given that in the examples analysed in this paper the Gini impurity signatures and the Shannon entropy signatures produced very similar results, the larger familiarity of a general physics audience with the latter influenced us to choose it over the former. Nevertheless, the question of which split criterion  should be used when training decision trees is as of yet largely undecided \cite{raileanu2004theoretical} and may be of relevance in the analysis of other physical systems than the ones studied here.

@@

\begin{figure}
\centering
\subfigure[]{\label{ssh1_heatmap_0}\includegraphics[width=.32\textwidth]{./ssh1/real_space_all_sites/imshow_winding_grid_winding_0_sim.png}}
\subfigure[]{\label{ssh1_heatmap_1}\includegraphics[width=.32\textwidth]{./ssh1/real_space_all_sites/imshow_winding_grid_winding_1_sim.png}}
\subfigure[]{\label{ssh1_heatmap}\includegraphics[width=.32\textwidth]{./ssh1/real_space_all_sites/merge_imshow_winding_grids_second_sim.png}}
\caption{Probability heatmaps learned by a combination of decision trees with eigenvector ensembling from bulk real space eigenvector data in experiment 1 (section \ref{experiment_1}). Heatmaps were averaged across all 100 iterations of the experiment. (a) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 0.  (b) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 1. (c) The phase diagram resulting from heatmaps (a) and (b).}
\label{ssh1_heatmaps}
\end{figure}

\begin{figure}
\centering
\subfigure[]{\label{ssh2_heatmap_-1}\includegraphics[width=.32\textwidth]{./ssh2/real_space_all_sites/imshow_winding_grid_winding_-1_sim.png}}
\subfigure[]{\label{ssh2_heatmap_0}\includegraphics[width=.32\textwidth]{./ssh2/real_space_all_sites/imshow_winding_grid_winding_0_sim.png}}
\subfigure[]{\label{ssh2_heatmap_1}\includegraphics[width=.32\textwidth]{./ssh2/real_space_all_sites/imshow_winding_grid_winding_1_sim.png}}
\subfigure[]{\label{ssh2_heatmap_2}\includegraphics[width=.32\textwidth]{./ssh2/real_space_all_sites/imshow_winding_grid_winding_2_sim.png}}
\subfigure[]{\label{ssh2_heatmap}\includegraphics[width=.32\textwidth]{./ssh2/real_space_all_sites/merge_imshow_winding_grids_sim.png}}
\caption{Probability heatmaps learned by a combination of random forests with eigenvector ensembling from bulk real space eigenvector data in experiment 2 (section \ref{experiment_2}). Heatmaps were averaged across all 100 iterations of the experiment. (a) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to -1.  (b) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 0. (c) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 1. (d) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 2. (e) The phase diagram resulting from heatmaps (a)-(d).}
\label{ssh2_heatmaps}
\end{figure}

@@

In this section, we explore the information entropy signatures obtained in section \ref{information_entropy_signatures} from a signal processing perspective and show how the same machinery used previously is able to perform two important unsupervised learning tasks on topological systems: dimensionality reduction and data compression.

As we have argued in section \ref{information_entropy_signatures}, the information entropy signatures can be used as a powerful dimensionality reduction tool: by keeping only the most relevant lattice sites in an information entropy signature, the amount of data needed to characterize the topological phase of a SSH system can be largely decreased with little loss of information. Nevertheless, given that the topological phase of a SSH system is a global property of the whole lattice, it is natural to expect that it should be possible to engineer global features from real space coordinates. Here we show how the symmetries in real space eigenvectors and information entropy signatures can be exploited to engineer new global topological features, leading to an effective compression of topological information.

@@

The topological features in both equations \eqref{DCT} and \eqref{DST} are generated from only half of the real space lattice, i.e. the sites $0 \leq l \leq \frac{N}{2}$. This is due to the fact that each equation assumes that the eigenvectors are even-symmetric (DCT topological features) or odd-symmetric (DST topological features) around the lattice sites $0$ and $\frac{N}{2}$ in real space as well. While these assumptions are strictly true for the eigenvector representations in wavevector space, this is not generally true for the real space representations. Therefore, equations \eqref{DCT} and \eqref{DST} achieve lattice compression by keeping only half of the real space eigenvector coordinates and imposing the corresponding boundary conditions (even symmetry or odd symmetry) on the lattice sites $0$ and $\frac{N}{2}$ to extrapolate the information from half of the lattice in real space.

@@

In strategy \emph{i)} the lattice sites used in the SSH 1 and SSH 2 systems were
\begin{equation}
\label{lattice_subsets}
\mathcal{S}_1 = (0, 1, 3, 50 ) \quad \text{and} \quad  \mathcal{S}_2 =(0, 1, 2, 3, 4, 5, 6, 7, 46, 48, 49, 50).   
\end{equation}
Note that for the nearest-neighbor SSH systems $\mathcal{S}_1$ corresponds to the four most informative sites such that $0\leq j \leq 50$ as indicated by the information entropy signature in figure \ref{feature_importances_ssh1}. Similarly, for the first- and second- nearest neighbors SSH systems $\mathcal{S}_2$ corresponds to the twelve most informative sites such that $0\leq j \leq 50$ in the corresponding information entropy signature in figure \ref{feature_importances_ssh2}. In table \ref{accuracy_scores}, the features used in strategy \emph{i)} are referred to as \xSOne and \xSTwo, according to the SSH systems they relate to.

@@

Similarly to strategy \emph{i)}, in strategy \emph{iii)} we selected the DCT topological features used in the SSH 1 and SSH 2 systems,
\begin{equation}
\label{DCT_subsets}
\mathcal{E}_1 = (1, 2, 36, 49) \quad \text{and} \quad  \mathcal{E}_2 =(0, 1, 2, 3, 4, 5, 6, 7, 47, 48, 49, 50) \qquad \text{(DCT)}  
\end{equation}
and the DST topological features,
\begin{equation}
\label{DST_subsets}
\mathcal{O}_1 =  (0, 18, 28, 30) \quad \text{and} \quad  \mathcal{O}_2 = (0, 1, 2, 3, 4, 5, 43, 44, 45, 46, 47, 48) \qquad \text{(DST)}.  
\end{equation}
The features in strategy \emph{iii)} are denoted in table \ref{accuracy_scores} by \xcEOne, \xsOOne, \xcETwo, \xsOTwo,   according to the topological features and wavevector space components used with each SSH systems.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Defining figures directory
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Figures for Main article
%%% phase diagrams (figure 1)
\newcommand\phaseDiagramsFigs{./00_article/phase_diagrams/}
\newcommand\SSHOnePhaseDiagram{\phaseDiagramsFigs ssh1.png}
\newcommand\SSHTwoPhaseDiagram{\phaseDiagramsFigs ssh2.png}
%%% ssh1 experiments (figure 2)
\newcommand\SSHOneExperimentsFigs{./00_article/ssh1/real_space_all_sites/}
\newcommand\SSHOneTrainValTestSplit{\SSHOneExperimentsFigs scatter_train_val_test_experiment_0.png}
\newcommand\SSHOneWindingTrain{\SSHOneExperimentsFigs scatter_winding_train_experiment_0.png}
\newcommand\SSHOnePredictionGrid{\SSHOneExperimentsFigs pcolormesh_prediction_grid_experiment_0.png}
%%% ssh2 experiments (figure 3)
\newcommand\SSHTwoExperimentsFigs{./00_article/ssh2/real_space_all_sites/}
\newcommand\SSHTwoTrainValTestSplit{\SSHTwoExperimentsFigs scatter_train_val_test_experiment_0.png}
\newcommand\SSHTwoWindingTrain{\SSHTwoExperimentsFigs scatter_winding_train_experiment_0.png}
\newcommand\SSHTwoPredictionGrid{\SSHTwoExperimentsFigs pcolormesh_prediction_grid_experiment_0.png}
%%% ssh1 probability heatmaps  (figure 4)
\newcommand\SSHOneWindingZero{\SSHOneExperimentsFigs imshow_winding_grid_winding_0_sim.png}
\newcommand\SSHOneWindingOne{\SSHOneExperimentsFigs imshow_winding_grid_winding_1_sim.png}
\newcommand\SSHOneHeatmap{\SSHOneExperimentsFigs merge_imshow_winding_grids_second_sim.png}
%%% ssh2 probability heatmaps  (figure 5)
\newcommand\SSHTwoWindingMinusOne{\SSHTwoExperimentsFigs imshow_winding_grid_winding_-1_sim.png}
\newcommand\SSHTwoWindingZero{\SSHTwoExperimentsFigs imshow_winding_grid_winding_0_sim.png}
\newcommand\SSHTwoWindingOne{\SSHTwoExperimentsFigs imshow_winding_grid_winding_1_sim.png}
\newcommand\SSHTwoWindingTwo{\SSHTwoExperimentsFigs imshow_winding_grid_winding_2_sim.png}
\newcommand\SSHTwoHeatmap{\SSHTwoExperimentsFigs merge_imshow_winding_grids_sim.png}
%%% information entropy signatures (figure 6)
\newcommand\SSHOneEntropySignature{\SSHOneExperimentsFigs plot_feature_importances.png}
\newcommand\SSHTwoEntropySignature{\SSHTwoExperimentsFigs plot_feature_importances.png}

%%%%%%%%%%%%%%%%%%%%%%%%%%%% Figures for Supplementary material
\newcommand\SuppFigures{./01_supplementary/}
\newcommand\SSH{\SuppFigures ssh/}
\newcommand\SSHOneSupp{\SuppFigures ssh1/}
%%% lattice size 100
\newcommand\SSHOneSuppOneZeroZero{\SSHOneSupp periodic_100_6561/}
\newcommand\SSHOneSuppOneZeroZeroCompressedRealSpace{\SSHOneSuppOneZeroZero real_space_best_04_sites/}
\newcommand\SSHOneSuppOneZeroZeroCompressedDCT{\SSHOneSuppOneZeroZero dct_best_04_wavevector_sites_from_best_04_real_space_sites/}
\newcommand\SSHOneSuppOneZeroZeroCompressedDST{\SSHOneSuppOneZeroZero dst_best_04_wavevector_sites_from_best_04_real_space_sites/}
%%% lattice size 140
\newcommand\SSHOneSuppOneFourZero{\SSHOneSupp periodic_140_6561/}
%\newcommand\SSHOneSuppOneFourZeroCompressedRealSpace{\SSHOneSuppOneFourZero real_space_best_04_sites/}
%\newcommand\SSHOneSuppOneFourZeroCompressedDCT{\SSHOneSuppOneFourZero dct_best_04_wavevector_sites_from_best_04_real_space_sites/}
%\newcommand\SSHOneSuppOneFourZeroCompressedDST{\SSHOneSuppOneFourZero dst_best_04_wavevector_sites_from_best_04_real_space_sites/}

%%%% lattice size 180
\newcommand\SSHOneSuppOneEightZero{\SSHOneSupp periodic_180_6561/}

\newcommand\SSHOneSuppTwoTwoZero{\SSHOneSupp periodic_220_6561/}
\newcommand\SSHTwoSupp{\SuppFigures ssh2/}
\newcommand\SSHTwoSuppOneZeroZero{\SSHTwoSupp periodic_100_6561/}
\newcommand\SSHTwoSuppOneZeroZeroCompressedRealSpace{\SSHTwoSuppOneZeroZero real_space_best_12_sites/}
\newcommand\SSHTwoSuppOneZeroZeroCompressedDCT{\SSHTwoSuppOneZeroZero dct_best_12_wavevector_sites_from_best_12_real_space_sites/}
\newcommand\SSHTwoSuppOneZeroZeroCompressedDST{\SSHTwoSuppOneZeroZero dst_best_12_wavevector_sites_from_best_12_real_space_sites/}

\newcommand\SSHTwoSuppOneFourZero{\SSHTwoSupp periodic_140_6561/}
\newcommand\SSHTwoSuppOneEightZero{\SSHTwoSupp periodic_180_6561/}
\newcommand\SSHTwoSuppTwoTwoZero{\SSHTwoSupp periodic_220_6561/}
%%% SSH model (figure 7)
\newcommand\SSHModel{\SSH ssh_model.pdf}
%%% Winding number (figure 8)
\newcommand\WindingNumber{\SSH winding.png}
%%% SSH1 compressed phase diagrams (figure 9)
\newcommand\SSHOneCompressedRealSpacePhaseDiagram{\SSHOneSuppOneZeroZeroCompressedRealSpace merge_imshow_winding_grids_second_sim.png}
\newcommand\SSHOneCompressedDCTPhaseDiagram{\SSHOneSuppOneZeroZeroCompressedDCT merge_imshow_winding_grids_second_sim.png}
\newcommand\SSHOneCompressedDSTPhaseDiagram{\SSHOneSuppOneZeroZeroCompressedDST merge_imshow_winding_grids_second_sim.png}
%%% SSH2 compressed phase diagrams (figure 9)
\newcommand\SSHTwoCompressedRealSpacePhaseDiagram{\SSHTwoSuppOneZeroZeroCompressedRealSpace merge_imshow_winding_grids_sim.png}
\newcommand\SSHTwoCompressedDCTPhaseDiagram{\SSHTwoSuppOneZeroZeroCompressedDCT merge_imshow_winding_grids_sim.png}
\newcommand\SSHTwoCompressedDSTPhaseDiagram{\SSHTwoSuppOneZeroZeroCompressedDST merge_imshow_winding_grids_sim.png}
%%% SSH1 entropy signatures on longer lattices (figure 10)
\newcommand\SSHOneEntropySignatureOneFourZero{\SSHOneSuppOneFourZero plot_feature_importances.png}
\newcommand\SSHOneEntropySignatureOneEightZero{\SSHOneSuppOneEightZero plot_feature_importances.png}
\newcommand\SSHOneEntropySignatureTwoTwoZero{\SSHOneSuppTwoTwoZero plot_feature_importances.png}
%%% SSH2 entropy signatures on longer lattices (figure 11)
\newcommand\SSHTwoEntropySignatureOneFourZero{\SSHTwoSuppOneFourZero plot_feature_importances.png}
\newcommand\SSHTwoEntropySignatureOneEightZero{\SSHTwoSuppOneEightZero plot_feature_importances.png}
\newcommand\SSHTwoEntropySignatureTwoTwoZero{\SSHTwoSuppTwoTwoZero plot_feature_importances.png}
%%% SSH1 cumulative information entropy signatures (figure 12)
\newcommand\SSHOneCumulativeEntropyOneFourZero{\SSHOneSuppOneFourZero plot_cumulative_feature_importances.png}
\newcommand\SSHOneCumulativeEntropyOneEightZero{\SSHOneSuppOneEightZero plot_cumulative_feature_importances.png}
\newcommand\SSHOneCumulativeEntropyTwoTwoZero{\SSHOneSuppTwoTwoZero plot_cumulative_feature_importances.png}
%%% SSH2 cumulative information entropy signatures (figure 12)
\newcommand\SSHTwoCumulativeEntropyOneFourZero{\SSHTwoSuppOneFourZero plot_cumulative_feature_importances.png}
\newcommand\SSHTwoCumulativeEntropyOneEightZero{\SSHTwoSuppOneEightZero plot_cumulative_feature_importances.png}
\newcommand\SSHTwoCumulativeEntropyTwoTwoZero{\SSHTwoSuppTwoTwoZero plot_cumulative_feature_importances.png}

%\definecolor{antiquefuchsia}{rgb}{0.57, 0.36, 0.51}

@@

\begin{abstract}
We develop a supervised machine learning algorithm that is able to learn topological phases of finite condensed matter systems from bulk data in real lattice space. The algorithm employs diagonalization in real space together with any supervised learning algorithm to learn topological phases through an eigenvector ensembling procedure. We combine our algorithm with decision trees and random forests to successfully recover topological phase diagrams of Su-Schrieffer-Heeger (SSH) models from bulk lattice data in real space and show how the Shannon information entropy of ensembles of lattice eigenvectors can be used to retrieve a signal detailing how topological information is distributed in the bulk. We further use insights obtained from these information entropy signatures to engineer global topological features from real space lattice data that still carry most of the topological information in the lattice, while greatly diminishing the size of feature space, thus effectively amounting to a topological lattice compression. Finally, we explore the theoretical possibility of interpreting the information entropy topological signatures in terms of emergent information entropy wave functions, which lead us to Heisenberg and Hirschman uncertainty relations for topological phase transitions. The discovery of Shannon information entropy signals associated with topological phase transitions from the analysis of data from several thousand SSH systems illustrates how model explainability in machine learning can advance the research of exotic quantum materials with properties that may power future technological applications such as qubit engineering for quantum computing. 
\end{abstract}

@@

The advantage of using decision tree-based algorithms to learn topological phases from local eigenvector data is that their use of entropy-based cost functions (such as Shannon information entropy or Gini impurity) furnishes them with an intrinsinc model explanaibility tool that summarizes how important each feature was to learning the desired patterns in the data. This makes it much easier to trace the localization of relevant information along the features of a data set. Here we use the Shannon information entropy of ensembles of real space eigenvectors to recover a signal quantifying the amount of topological information available from each lattice site. This is a highly non-trivial proposition since the topological phase of a system is a global property of the system as a whole emerging from complex interactions between its components, and therefore even defining a local topological signal is a daunting theoretical task. To our knowledge this is the first time that a signal describing the localization of topological information in the bulk of topological condensed matter systems is presented in the literature. These topological signals, here called information entropy signatures, are the subject of section \ref{information_entropy_signatures}.

@@

As of yet model explainability \cite{gilpin2018explaining}-\cite{roscher2020explainable} is one of the topics at the edge of machine learning research that has been little explored by the physics community working at the interface between the two disciplines. This raises important questions as to whether machine learning can in fact help to advance theoretical investigation in physics, since the majority of physics papers published on the subject are proofs of concept aimed at showing that modern machine learning techniques are capable of recognizing the relevant patterns in data from physical systems whose properties were known in advance. By proposing new concepts from the data analysis of physical systems of contemporary interest and knitting together ideas from topological phase transitions and information theory by dint of model explainability, we expect to draw the physics community's attention to this essential machine learning tool. 
\begin{figure}
\centering
\subfigure[]{\label{ssh1}\includegraphics[width=.45\textwidth]{\SSHOnePhaseDiagram}}\quad
\subfigure[]{\label{ssh2}\includegraphics[width=.45\textwidth]{\SSHTwoPhaseDiagram}}
\caption{Phase diagrams in parameter space. a) SSH model with first-neighbor hoppings $t_1$ and $t_2$. The (red) regions with winding number $W$ = 0 are trivial, while the (blue) regions with winding number $W$ = 1 are topologically non-trivial. b) SSH model with first ($t_1$ and $t_2$) and second ($T_1$ and $T_2$) nearest-neighbor hoppings. In this article we set $t_1$ = $t_2$ = 1 and renamed the variables $T_1$ $\rightarrow$ $t_1$, $T_2$ $\rightarrow$ $t_2$ for convenience. The (orange) region with winding number $W$ = 0 is trivial while the others with winding numbers $W$ = -1, $W$ = 1 and $W$ = 2 (red, green and blue respectively) are topologically non-trivial.}
\label{fig:phasediagrams}
\end{figure}

@@

Figures \ref{figexp1_exp} and \ref{figexp2_exp} respectively illustrate single iterations of experiment 1 (section \ref{experiment_1}) and experiment 2 (section \ref{experiment_2}) as seen from parameter space. The accuracy statistics presented in the following subsections, as well as the probability heatmaps and recovered phase diagrams shown in figures \ref{ssh1_heatmaps} and \ref{ssh2_heatmaps} were obtained after bootstrapping each experiment $n_{exp}$ = \nExp\ times. Thus, each probability heatmap shown in figures \ref{ssh1_heatmap_0}, \ref{ssh1_heatmap_1} and \ref{ssh2_heatmap_-1} to \ref{ssh2_heatmap_2} represents the averaged fraction of eigenvectors of each Hamiltonian in the grid that were classified with a given winding number across 100 experiments. The recovered phase diagrams \ref{ssh1_heatmap} and \ref{ssh2_heatmap} are constructed by superposing the corresponding probability heatmaps. As these figure make clear, the recovered phase diagrams faithfully portray the true phase diagrams in figure \ref{fig:phasediagrams}, with clear phase transition lines appearing in the regions of highest uncertainty.

@@

\begin{figure}
\centering
\subfigure[]{\label{figexp1_exp:a}\includegraphics[width=.32\textwidth]{\SSHOneTrainValTestSplit}}
\subfigure[]{\label{figexp1_exp:b}\includegraphics[width=.32\textwidth]{\SSHOneWindingTrain}}
\subfigure[]{\label{figexp1_exp:c}\includegraphics[width=.32\textwidth]{\SSHOnePredictionGrid}}
\caption{Visualization of a single iteration of experiment 1 (section \ref{experiment_1}) as seen from 2D parameter space. (a) Train/validation/test split. (b) Distribution of winding numbers in the training set. (c) Phase diagram learned from real space lattice data by combining a decision tree with eigenvector ensembling.}
\label{figexp1_exp}
\end{figure}

@@

\begin{figure}
\centering
\subfigure[]{\label{figexp2_exp:a}\includegraphics[width=.32\textwidth]{\SSHTwoTrainValTestSplit}}
\subfigure[]{\label{figexp2_exp:b}\includegraphics[width=.32\textwidth]{\SSHTwoWindingTrain}}
\subfigure[]{\label{figexp2_exp:c}\includegraphics[width=.32\textwidth]{\SSHTwoPredictionGrid}}
\caption{Visualization of a single iteration of experiment 2 (section \ref{experiment_2}) as seen from 2D parameter space. (a) Train/validation/test split. (b) Distribution of winding numbers in the training set. (c) Phase diagram learned from real space lattice data by combining a random forest with eigenvector ensembling.}
\label{figexp2_exp}
\end{figure}

@@

We should briefly comment on the possibility of using the Gini impurity of ensembles of eigenvectors \cite{friedman2001elements} instead of their Shannon entropy as cost function. This would similarly lead us to Gini impurity signatures of topological phase transitions. Given that in the examples analysed in this paper the Gini impurity signatures and the Shannon entropy signatures were very similar, the larger familiarity of a general physics audience with the latter influenced us to choose it over the former. Nevertheless, the question of which split criterion  should be used when training decision trees is as of yet largely undecided \cite{raileanu2004theoretical} and may be of relevance in the analysis of other physical systems than the ones studied here.

@@

\begin{figure}
\centering
\subfigure[]{\label{ssh1_heatmap_0}\includegraphics[width=.32\textwidth]{\SSHOneWindingZero}}
\subfigure[]{\label{ssh1_heatmap_1}\includegraphics[width=.32\textwidth]{\SSHOneWindingOne}}
\subfigure[]{\label{ssh1_heatmap}\includegraphics[width=.32\textwidth]{\SSHOneHeatmap}}
\caption{Probability heatmaps learned by a combination of decision trees with eigenvector ensembling from bulk real space eigenvector data in experiment 1 (section \ref{experiment_1}). Heatmaps were averaged across all 100 iterations of the experiment. (a) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 0.  (b) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 1. (c) The phase diagram resulting from heatmaps (a) and (b).}
\label{ssh1_heatmaps}
\end{figure}

\begin{figure}
\centering
\subfigure[]{\label{ssh2_heatmap_-1}\includegraphics[width=.32\textwidth]{\SSHTwoWindingMinusOne}}
\subfigure[]{\label{ssh2_heatmap_0}\includegraphics[width=.32\textwidth]{\SSHTwoWindingZero}}
\subfigure[]{\label{ssh2_heatmap_1}\includegraphics[width=.32\textwidth]{\SSHTwoWindingOne}}
\subfigure[]{\label{ssh2_heatmap_2}\includegraphics[width=.32\textwidth]{\SSHTwoWindingTwo}}
\subfigure[]{\label{ssh2_heatmap}\includegraphics[width=.32\textwidth]{\SSHTwoHeatmap}}
\caption{Probability heatmaps learned by a combination of random forests with eigenvector ensembling from bulk real space eigenvector data in experiment 2 (section \ref{experiment_2}). Heatmaps were averaged across all 100 iterations of the experiment. (a) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to -1.  (b) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 0. (c) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 1. (d) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 2. (e) The phase diagram resulting from heatmaps (a)-(d).}
\label{ssh2_heatmaps}
\end{figure}

@@

In this section, we explore the information entropy signatures obtained in section \ref{information_entropy_signatures} from a signal processing perspective. As we shall see, this will enable us to perform two important unsupervised learning tasks on topological systems: dimensionality reduction and data compression.

As a cursory glance at figures \ref{feature_importances_ssh1} and \ref{feature_importances_ssh2} seems to suggest, the existence of peaks and symmetries in information entropy signatures can be exploited as a powerful dimensionality reduction tool: by keeping only the most relevant lattice sites in an information entropy signature, the amount of data needed to characterize the topological phase of a SSH system can be largely decreased with little loss of information. Nevertheless, given that the topological phase of a SSH system is a global property of the whole lattice, it is natural to expect that it should be possible to engineer global features from real space coordinates. Here we show how the symmetries in real space eigenvectors and information entropy signatures can be exploited to engineer new global topological features, leading to an effective compression of topological information.

@@

The topological features in both equations \eqref{DCT} and \eqref{DST} are generated from only half of the real space lattice, i.e. the sites $0 \leq l \leq \frac{N}{2}$. This is due to the fact that each equation assumes that the eigenvectors are even-symmetric (DCT topological features) or odd-symmetric (DST topological features) around the lattice sites $0$ and $\frac{N}{2}$ in real space as well. While these assumptions are strictly true for the eigenvector representations in wavevector space, they are not generally true for the real space representations. Therefore, equations \eqref{DCT} and \eqref{DST} achieve lattice compression by keeping only half of the real space eigenvector coordinates and imposing the corresponding boundary conditions (even symmetry or odd symmetry) on the lattice sites $0$ and $\frac{N}{2}$ to extrapolate the information from one half of the lattice to the other.

@@

In strategy \emph{i)} the lattice sites used in the SSH 1 and SSH 2 systems were
\begin{equation}
\label{lattice_subsets}
\mathcal{S}_1 = (0, 1, 3, 50 ) \quad \text{and} \quad  \mathcal{S}_2 =(0, 1, 2, 3, 4, 5, 6, 7, 46, 48, 49, 50).   
\end{equation}
Note that for the nearest-neighbor SSH systems $\mathcal{S}_1$ corresponds to the four most informative sites such that $0\leq l \leq 50$ as indicated by the information entropy signature in figure \ref{feature_importances_ssh1}. Similarly, for the first- and second- nearest neighbors SSH systems $\mathcal{S}_2$ corresponds to the twelve most informative sites such that $0\leq l \leq 50$ in the corresponding information entropy signature in figure \ref{feature_importances_ssh2}. In table \ref{accuracy_scores}, the features used in strategy \emph{i)} are referred to as \xSOne and \xSTwo, according to the SSH systems they relate to.

@@

Similarly to strategy \emph{i)}, in strategy \emph{iii)} we selected the most informative DCT topological features of both SSH 1 and SSH 2 systems that were obtained from strategy \emph{ii)},
\begin{equation}
\label{DCT_subsets}
\mathcal{E}_1 = (1, 2, 36, 49) \quad \text{and} \quad  \mathcal{E}_2 =(0, 1, 2, 3, 4, 5, 6, 7, 47, 48, 49, 50) \qquad \text{(DCT)}  
\end{equation}
and the most informative DST topological features of both systems that were obtained from strategy \emph{ii)} as well,
\begin{equation}
\label{DST_subsets}
\mathcal{O}_1 =  (0, 18, 28, 30) \quad \text{and} \quad  \mathcal{O}_2 = (0, 1, 2, 3, 4, 5, 43, 44, 45, 46, 47, 48) \qquad \text{(DST)}.  
\end{equation}
The features in strategy \emph{iii)} are denoted in table \ref{accuracy_scores} by \xcEOne, \xsOOne, \xcETwo, \xsOTwo,   according to the topological features and wavevector space subset used with each SSH system.
